WASP-4 is Accelerating Towards the Earth

### 2020-01-28 paper/ms.tex
update the paper

% TODO:
%
% * email to full collaborator list
% * add any missing CPS coauthors
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

% TODO: may want to improve title
\title{WASP-4 is Accelerating Towards the Earth}
% ": A Cautionary Tale in the Search for Tidal Orbital Decay"
% "An Explanation for WASP-4's Decreasing Orbital Period".
% "WASP-4's Decreasing Orbital Period: Solved."

@@

%
% key authors: did Rachel do the Zorro reductions?
%
\author[0000-0002-0514-5538]{L. G. Bouma}
\affiliation{ Department of Astrophysical Sciences, Princeton
University, 4 Ivy Lane, Princeton, NJ 08540, USA}
%
\author[0000-0002-4265-047X]{J. N. Winn}
\affiliation{ Department of Astrophysical Sciences, Princeton
University, 4 Ivy Lane, Princeton, NJ 08540, USA}

%
% contributing authors: alphabetical
%
\author[0000-0002-0531-1073]{H. Isaacson}
\affiliation{Astronomy Department, University of California, Berkeley,
CA 94720, USA}
%
\author[0000-0001-8638-0320]{A. W. Howard}
\affiliation{Cahill Center for Astrophysics, California Institute of
Technology, Pasadena, CA 91125, USA}
%
\author{S. B. Howell}
\affiliation{NASA Ames Research Center, Moffett Field, CA 94035, USA}
%
\author{H. Knutson}
\affiliation{Division of Geological and Planetary Sciences, California
Institute of Technology, Pasadena, CA 91125, USA}
%
%
%\author{R. A. Matson}
% \affiliation{NASA Ames Research Center, Moffett Field, CA 94035, USA}
%

\begin{abstract}
  Space and ground-based transit observations of the hot Jupiter
  WASP-4b have recently shown that its orbital period is decreasing.
  Potential causes of the period change include tidal orbital decay,
  orbital precession, and light-travel time effects.
  In this work we present new radial velocity measurements of WASP-4,
  acquired with Keck-HIRES.
  The new data show that the system is accelerating towards the Earth
  at $\dot{\gamma} = -0.0422^{+0.0028}_{-0.0027}\ {\rm m}\,{\rm
  s}^{-1}\,{\rm yr}^{-1}$.
  The corresponding period decrease explains WASP-4's
  changing orbital period as a light-travel time effect.
  Combining the radial velocities with new speckle imaging limits, we
  find that system's acceleration is likely caused by a
  10-200$\,M_{\rm Jup}$ companion with semi-major axis between 10-100$\,$AU.
  Based on statistics from \citet{knutson_friends_2014}, about 1 in 6
  hot Jupiters are expected to show period decreases commensurate with
  WASP-4 due to acceleration by outer companions.
  Continued long-term radial velocity monitoring of hot Jupiters is
  therefore essential to distinguish tidal orbital decay from
  line-of-sight accelerations.
\end{abstract}

@@

The main focus of this study is the hot Jupiter WASP-4b, which has an
orbital period that from transits appears to be decreasing by about 10
milliseconds per year.  We discovered the timing variations using
data from TESS and a decade of ground-based monitoring
\citep[][hereafter \citetalias{bouma_wasp4b_2019}]{bouma_wasp4b_2019}.
Thereafter, \citet{southworth_transit_2019} reported 22 new transit
times for the system, and confirmed that the updated series of transit
times was consistent with a quadratic ephemeris.  With additional
data, they found a lower best-fit decay rate of $\dot{P} = -XX.X \pm
X.X$ milliseconds per year.
A separate study by \citet{baluev_2019} reported
additional archival transit light curves of WASP-4b, most notably from
the TRAPPIST archive, and a select few other observers.

To determine the origin of the period change, we acquired four
additional radial velocity measurements using Keck-HIRES, extending
the RV baseline from 3 to 9 years.  Previously, the radial velocities
were sparse and showed a weak ($<2\sigma$) linear trend.  Our new
measurements reveal a line-of-sight acceleration of $\dot{\gamma} =
-0.0422^{+0.0028}_{-0.0027}\ {\rm m\,s^{-1}\,yr^{-1}}$.  This
translates to an expected period
decrease simply from the light travel time effect of ZZ milliseconds
per year --- about commensurate with what is observed.
In the following, we avoid the term ``R{\o}mer delay'' as this phrase
typically signifies arrival time delay due to observatory motion. We
are discussing an analog of the Doppler effect for a source with
constant line-of-sight acceleration.

% FIXME: figure out correct sections
\S~\ref{sec:observations} presents the new observations...
\S~\ref{sec:analysis} describes our analysis...
\S~\ref{sec:discussion} discusses...
\S~\ref{sec:conclusions} gives conclusions...

@@

{\bf Table~N} lists the transit times we used in our analysis.  We
include data from the peer-reviewed literature for which (a) the
analysis was based on observations of a single transit, (b) the
midpoint was fitted as a free parameter, and (c) the time system
specified both the leap second correction (TDB or UTC) and also
whether any barycentric or heliocentric corrections had been
performed.

% FIXME : verify with Michael Gillon that this is true. (P. Evans
% done)
The majority of times are identical to those we collected in
\citetalias{bouma_wasp4b_2019}.  Twenty-two new times reported by
\citet{southworth_transit_2019} are included.  These transits were
observed from the 3.58m NTT and Danish 1.54m telescopes at La Silla,
and the SAAO 1.0m telescope.  Additional timimg
measurements have also recently been reported by \citet{baluev_2019},
based on a homogeneous analysis of archival ground-based observations
taken professional and amateur observers.  We included
twelve of their ``high quality'' transit times acquired by TRAPPIST
(six transits), El~Sauce (four transits), and
\citet{petrucci_no_2013}.  For TRAPPIST and El~Sauce, we verified with
the original observers that correct barycentric and leap-second
corrections had been performed (M.~Gillon, P.~Evans, priv{.} comm{.}).
We omitted the fourteen remaining \citeauthor{baluev_2019}
ETD\footnote{\url{http://var2.astro.cz/ETD}} times due to ambiguity in
whether leap-second corrections had or had not been performed.

@@

A cursory analysis of the new HIRES observations led to our detection
of a linear trend in the residuals after fitting out the orbit of
WASP-4b.  This prompted us to acquire speckle imaging using Zorro at
Gemini-South \citep[see][, and the instrument
web-pages\footnote{\url{www.gemini.edu/sciops/instruments/alopeke-zorro/}}]{scott_nessi_2018}.
Zorro (and its counterpart, 'Alopeke, on Gemini-North) are
dual-channel speckle interferometers that can be used to observe in
narrow-band filters centered at 562$\,$nm and 832$\,$nm.

We observed WASP-4 twice, on the night of September 11-12 (8 sets of
1000 frames each) and also on the night of September 28-29 (7 sets of
1000 frames each). The second night had better seeing and produced a
slightly better result, which we opt to use for our analysis.

% TODO: might Howell suggest a paragraph for how the contrast ratios
% were derived?
{\bf We reduced the speckle images to contrast curves by ...}

@@

We considered two possible models for the observed transit times.  The
first model assumes a constant orbital period on a circular orbit:
\begin{align}
  t_{\rm tra}(E) &= t_0 + PE,
\end{align}
where $E$ is the epoch number and $t_0$ is a reference epoch.  The
second model assumes the period changes at a steady rate:
\begin{align}
  t_{\rm tra}(E) &=
    t_0 + PE +
    \frac{1}{2} \frac{{\rm d}P}{{\rm d}E} E^2.
\end{align}
The free parameters are the reference epoch $t_0$, the period at the
reference epoch, and the period derivative, ${\rm d}P/{\rm d}t = (1/P)
{\rm d}P/{\rm d}E$.  We defined the epoch numbers such that $E=0$ is
near the weighted average of the observed times.  This helps to reduce
the covariance between $t_0$ and $P$.  A third possible model that we
did not consider for reasons that will become apparent is a
precessing, eccentric orbit \citep[{\it
e.g.},][]{gimenez_revision_1995,patra_2017}.

@@

Figure~\ref{fig:times} shows the observed transit times, minus the
best-fit constant period model.  The best-fitting constant-period
model has $\chi^2 = 276$, 91 degrees of freedom, and $\chi^2_{\rm
red}=3.0$.  The best-fitting quadratic model has $\chi^2 = 183$, 90
degrees of freedom, and $\chi^2_{\rm red}=2.0$.  From the reduced
$\chi^2$ values, we can surmise that neither model entirely describes
the transit data that we have collected---there must be some
additional source of signal or potentially noise. In
\citet{bouma_wasp4b_2019}, we found that the quadratic model for the
(less complete) transit data gave $\chi^2_{\rm red}=1.0$.  The only
difference between our old and new transit datasets are the addition
of transit times from \citet{southworth_transit_2019} and
\citet{baluev_2019}.  The worsened $\chi^2_{\rm red}$ could reflect
underestimated statistical uncertainties in these or in fact any of
the transit measurements. It could also reflect systematic errors in
the time-systems in which the transit measurements were recorded,
though we have taken every caution against the latter possibility.
% For purposes of simplicity, we simply accept that the additional data
% have effectively increased the scatter in the transit O-C diagram.
The difference in the BIC \citep{kass_bayes_1995} between the linear
and quadratic and models is $\Delta {\rm BIC} = 89.1$, strongly
favoring the quadratic case.

@@

%FIXME: perhaps include the WASP-4b phase-fold here too??
\begin{figure}[t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.5\textwidth]{f2.pdf}
	\end{center}
	\vspace{-0.7cm}
	\caption{
    {\bf Radial velocity observations of WASP-4.} The best-fit Keplerian orbit
    of WASP-4b has been subtracted.  The linear trend inferred from the
    RV data is shown with a black line, with $1\sigma$ errors in gray.
    The trend that would be needed to produce the period decrease seen
    in transits ($\dot{P}= - 8.64 \pm 0.89~{\rm ms}\,{\rm yr}^{-1}$) is
    indicated with the purple dotted line.  The four new RV
    measurements presented in this work increased the significance of
    the radial velocity trend from $\approx$2$\sigma$ to 15$\sigma$.
	\label{fig:rvs}
  \vspace{-0.3cm}
	}
\end{figure}

%FIXME: this entire paragraph copies Bouma+19 too heavily.
We began by fitting a single Keplerian orbit, plus instrument offsets,
jitters, and a long-term trend
\citep[][\texttt{radvel}]{fulton_radvel_2018}.  We set Gaussian priors
on the period and time of inferior conjunction using the values from
Table~4 of \citetalias{bouma_wasp4b_2019}, and fixed the eccentricity
to zero, per the results of \citet{beerer_secondary_2011},
\citet{knutson_friends_2014} and \citet{bonomo_gaps_2017}.  The free
parameters were the velocity semi-amplitude, the instrument
zero-points, the instrument jitters (an additive white noise term for
each instrument), linear ($\dot{v_{\rm r}}$), and optionally
second-order ($\ddot{v_{\rm r}}$) acceleration terms.

@@

WASP-4 was found to be accelerating towards our line-of-sight at high
confidence,
\begin{equation}
  \dot{v}_{\rm r} = \dot{\gamma} = 
     -0.0422^{+0.0028}_{-0.0027}
     \,{\rm m}\,{\rm s}^{-1}\,{\rm day}^{-1}.
\end{equation}
For comparison, before our recent observing run, 
$\dot{v}_{\rm r}$ was thought to be about five times smaller, and was
only marginally significant
\citep{knutson_friends_2014,bouma_wasp4b_2019}.

The system's acceleration towards our line-of-sight causes a decrease
in the apparent orbital period.  The period
derivative expected from radial velocities is
\begin{align}
  \dot{P}_{\rm RV} &= \frac{\dot{v}_{\rm r} P}{c} \\
  \dot{P}_{\rm RV} &= -5.94 \pm 0.39~{\rm ms}\,{\rm yr}^{-1}.
\end{align}
In other words, the majority of the period decrease observed in
transits ($\dot{P}= - 8.64 \pm 0.89~{\rm ms}\,{\rm yr}^{-1}$) seems to
be explained by the acceleration of the host star.

An important consideration is whether the measured RV trend is at all
correlated with stellar activity.  We investigate this by analyzing
WASP-4's emission in the Ca II H \& K lines, as quantified with the
chromospheric $S$-index \citep{wright_chromospheric_2004}.  We only
examined the HIRES velocities for this step, since they are the main
source of signal for our analysis.  First, we subtracted the orbital
solution from the Keck-HIRES velocities.  Then, following
\citet{bryan_statistics_2016,bryan_excess_2019}, we calculated the
Spearman rank correlation coefficient between the $S$-index and the
orbit-subtracted velocities.  We found a correlation coefficient of
0.16. Though suggestive, it is not statistically significant (the
corresponding $p$-value is 0.65).  Furthermore, inspection of the
$S$-index timeseries showed no secular or sinusoidal trends, as would
be expected if we were observing a long-term magnetic activity cycle.
The $S$-index values are included in {\bf Table X}.  We conclude that
it would be highly unlikely for the linear trend to be caused by
stellar activity.

@@

\begin{figure*}[t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.78\textwidth]{f4.pdf}
	\end{center}
	\vspace{-0.5cm}
	\caption{
    {\bf Masses and semi-major axes of companions that meet
    requirements of both the radial velocity and the speckle imaging.}
    The likelihood inferred from radial velocities is shown in
    grayscale, and the region excluded from the imaging is shown with
    a cross-hatch pattern.
	\label{fig:mass_sma}
  \vspace{-0.3cm}
	}
\end{figure*}

@@

Due to the custom filters of the Zorro imager, and corresponding lack
of synthetic photometry, we further assumed that all sources had
blackbody spectra. While this is clearly false, we do not readily have
access to the planetary and stellar atmosphere models needed for the
consistent calculation with the \texttt{COND03} and \texttt{MESA}
models.  We therefore adopted the effective temperatures and
bolometric luminosities from the \citet{baraffe_evolutionary_2003} and
MIST isochrones.  Using these theoretical quantities and the empircal
Zorro bandpass functions, we calculated absolute magnitudes in the 562
and 832 nm Zorro bandpasses for stellar and planetary mass companions.
Applying the same calculation to WASP-4 itself using the effective
temperature and bolometric luminosity from
\citetalias{bouma_wasp4b_2019}, we derived the transformation from
contrast ratio to companion mass.  The resulting limits are shown as
the cross-hatched region in Figure~\ref{fig:mass_sma}.

@@

%FIXME: may want to verify that changing the cutoff location, to say
%80Mjup or whatever the minimum BD mass is doesn't drastically alter
%the posterior
We assigned each companion a mass and semimajor axis from log-uniform
distributions within the grid cell. We drew the inclination from a
uniform distribution in $\cos i$.  For companion masses less than
$10\,M_{\rm Jup}$, we drew the eccentricity from
\citet{kipping_beta_2013}'s long-period exoplanet Beta distribution
($a=1.12$, $b=3.09$).  If the companion mass exceeded $10\,M_{\rm
Jup}$, we drew the eccentricity distribution from the power-law $p_e
\propto e^\eta$ reported by \citet{moe_mind_2017} in their
Equation~17.  We emphasize that the long-period exoplanet and
long-period binary distributions are quite different: the exoplanet
distribution is ``bottom-heavy'', with eccentricities preferentially
close to zero. The binary star distribution is ``top-heavy'', with
more eccentricities close to one.  We chose a mass cutoff of
$10\,M_{\rm Jup}$ based on the bound observed by
\citet{schlaufman_evidence_2018} between giant planets, presumably
formed through core accretion, and brown dwarfs, presumably formed via
gravitational instability.

For each simulated companion, we then drew a sample from the converged
chains of our initial model of WASP-4b, plus its linear trend. We
subtracted the planet's orbital solution, leaving the linear trend.
Given $(a_{\rm c}, M_{\rm c}, e_{\rm c}$ for each simulated outer
companion, and the fixed instrument offsets and jitters from the MCMC
chains, we then performed a maximum likelihood fit for the time and
argument of periastron of the outer simulated companion.  We converted
the resulting $128\times128\times512$ cube of best-fit log-likelihood
values to probabilities, and averaged over the samples in each grid
cell to derive a probability distribution in mass and semi-major axis.

The result is shown as grayscale background in
Figure~\ref{fig:mass_sma}.

@@

\subsection{Implications for WASP-4}
Previous possibilities suggested to explain the period decrease of
WASP-4b included tidal orbital decay, orbital precession, and
light-travel time effects \citep{bouma_wasp4b_2019}.  Our new radial
velocity measurements have now shown the least exotic
option---light-travel time effects---is also the most likely
explanation.  Transits show the obital period to decrease by $YY.Y \pm
Z.Z~{\rm ms}\,{\rm yr}^{-1}$; the line-of-sight acceleration observed
in radial velocities implies an expected period decrease of $YY.Y \pm
Z.Z~{\rm ms}\,{\rm yr}^{-1}$.  
Though the quantitative agreement is not perfect, Occam's razor would
suggest that the line of sight acceleration is probably a sufficient
explanation for the apparent decrease of WASP-4b's orbital period.
% While further radial velocity observations of the system should help
% in determining the mass and semi-major axis of the companion, it seems
% unlikely that additional transit observations will yield near-term
% constraints on orbital decay.

The corresponding requirements for the companion causing the
acceleration are that it is likely either a brown-dwarf or low mass
star, and that it is between $10$-$100\,{\rm AU}$ from the host star
(Figure~\ref{fig:mass_sma}).
Given such a mass, this companion could at one time have influenced
the orbital evolution of the inner giant.
Further radial velocity monitoring should eventually reveal the
orbital parameters and minimum mass of the companion.

% TODO: add a paragraph noting that while the skepticism
% is fine "I was right".

>>>>

@@

\title{WASP-4 is Accelerating Towards the Earth}
% ", Explaining its Decreasing Orbital Period"
% ": A Cautionary Tale in the Search for Tidal Orbital Decay"
% "An Explanation for WASP-4's Decreasing Orbital Period".
% "WASP-4's Decreasing Orbital Period: Solved."

@@

%
% key authors:
%
\author[0000-0002-0514-5538]{L. G. Bouma}
\affiliation{ Department of Astrophysical Sciences, Princeton
University, 4 Ivy Lane, Princeton, NJ 08540, USA}
%
\author[0000-0002-4265-047X]{J. N. Winn}
\affiliation{ Department of Astrophysical Sciences, Princeton
University, 4 Ivy Lane, Princeton, NJ 08540, USA}

%
% contributing authors: alphabetical
%
\author[0000-0001-8638-0320]{A. W. Howard}
\affiliation{Cahill Center for Astrophysics, California Institute of
Technology, Pasadena, CA 91125, USA}
%
\author{S. B. Howell}
\affiliation{NASA Ames Research Center, Moffett Field, CA 94035, USA}
%
\author[0000-0002-0531-1073]{H. Isaacson}
\affiliation{Astronomy Department, University of California, Berkeley,
CA 94720, USA}
%
\author{H. Knutson}
\affiliation{Division of Geological and Planetary Sciences, California
Institute of Technology, Pasadena, CA 91125, USA}
%
\author{R. A. Matson}
\affiliation{NASA Ames Research Center, Moffett Field, CA 94035, USA}
%

\begin{abstract}
  Space and ground-based transit observations of the hot Jupiter
  WASP-4b have recently shown that its orbital period is decreasing.
  The period change could be caused by tidal orbital decay, orbital
  precession, or light-travel time effects.  In this work we present
  new radial velocity measurements of WASP-4, acquired with
  Keck-HIRES.  The data show that the system is accelerating towards
  the Earth at $\dot{\gamma} = -0.0422^{+0.0028}_{-0.0027}\ {\rm
  m}\,{\rm s}^{-1}\,{\rm yr}^{-1}$.  This implied period decrease
  probably explains WASP-4's changing orbital period as a light-travel
  time effect.  By combining the radial velocities with new speckle
  imaging limits, we find that system's acceleration is likely caused
  by a 10-200$\,M_{\rm Jup}$ companion with semi-major axis between
  10-100$\,$AU.  Based on statistics from
  \citet{knutson_friends_2014}, about 1 in 6 hot Jupiters are expected
  to show period decreases commensurate with WASP-4 due to
  acceleration by outer companions.  Continued long-term radial
  velocity monitoring of hot Jupiters is therefore essential to
  distinguish tidal orbital decay from line-of-sight accelerations.
\end{abstract}

@@

The main focus of this study is the hot Jupiter WASP-4b, which has an
orbital period that appears to be decreasing by about 10 milliseconds
per year.  We discovered the timing variations using data from TESS
and a decade of ground-based observations \citep[][hereafter
\citetalias{bouma_wasp4b_2019}]{bouma_wasp4b_2019}.  Thereafter,
\citet{southworth_transit_2019} reported 22 new transit times for the
system, and found an updated decay rate of $\dot{P} = -9.2 \pm 1.1$
milliseconds per year. The \citeauthor{southworth_transit_2019} decay rate
was $\approx$3$\sigma$ less rapid than that found by
\citetalias{bouma_wasp4b_2019}, but the conclusions of the
studies were otherwise the same.  A separate study by
\citet{baluev_2019} reported additional archival transit light curves
of WASP-4b, most notably from the TRAPPIST archive, as well as from
select other observers.  \citet{baluev_2019} pointed out that
when using lower-precision subsets of the available transit data,
the case for a decreasing period worsened.  We agree with this
statement.

To determine the origin of the period change, we acquired four
additional radial velocity measurements using Keck-HIRES, extending
the RV baseline from 3 to 9 years.  Previously, the five available
HIRES radial velocities suggested a weak ($<2\sigma$) linear trend
\citep{knutson_friends_2014}.  Our new measurements reveal a
line-of-sight acceleration of $\dot{\gamma} =
-0.0422^{+0.0028}_{-0.0027}\ {\rm m\,s^{-1}\,yr^{-1}}$.  This
translates to an expected period decrease from the light-travel time
effect of -5.9 milliseconds per year---about commensurate with what
is observed from transits.  In the following, we avoid the term
``R{\o}mer delay'', which typically signifies arrival time
delay due to observatory motion. We are discussing the Doppler effect
seen by a stationary observer for a source with constant line-of-sight
acceleration.

In the following, \S~\ref{sec:observations} collects the available
transit data and presents the new radial velocity and speckle imaging
observations.  \S~\ref{sec:analysis} analyzes the data, and finds that
they yield a picture in which the WASP-4 system is accelerating
towards our line-of-sight, likely due to the pull of a brown or
M-dwarf companion.
\S~\ref{sec:discussion} places this result in the broader context of
orbital decay searches, and points out that line-of-sight
accelerations, {\it i.e.}, ``false positive orbital decay signals'',
are relatively common for the hot Jupiter population.
\S~\ref{sec:conclusions} offers brief concluding remarks.

@@

{\bf Table~N} lists the transit times we collected for our analysis.
We include data from the peer-reviewed literature for which (a) the
analysis was based on observations of a single transit, (b) the
midpoint was fitted as a free parameter, and (c) the time system
specified both the leap second correction (TDB or UTC) and also
whether any barycentric or heliocentric corrections had been
performed.

% FIXME : verify with Michael Gillon that this is true. (P. Evans
% done)
The majority of times are identical to those we collected in
\citetalias{bouma_wasp4b_2019}.  Twenty-two new times reported by
\citet{southworth_transit_2019} are included.  These transits were
observed from the 3.58m NTT and Danish 1.54m telescopes at La Silla,
and the SAAO 1.0m telescope.

Additional timimg measurements have also recently been reported by
\citet{baluev_2019}, based on a homogeneous analysis of archival
ground-based observations taken professional and amateur observers.
We included twelve of their ``high quality'' transit times acquired by
TRAPPIST (six transits), El~Sauce (four transits), and
\citet{petrucci_no_2013}.  For TRAPPIST and El~Sauce, we verified with
the original observers that correct barycentric and leap-second
corrections had been performed (M.~Gillon, P.~Evans, priv{.} comm{.}).
We omitted the fourteen remaining \citeauthor{baluev_2019}
ETD\footnote{\url{http://var2.astro.cz/ETD}} times due to ambiguity in
whether leap-second corrections had or had not been performed.

@@

A cursory analysis of the new HIRES observations led to our detection
of a linear trend in the residuals after fitting out the orbit of
WASP-4b.  This prompted us to acquire speckle images using Zorro at
Gemini-South \citep[see][and the instrument
web-pages\footnote{\url{www.gemini.edu/sciops/instruments/alopeke-zorro/}}]{scott_nessi_2018}.
Zorro is a dual-channel speckle interferometer with narrow-band
filters centered at 562$\,$nm and 832$\,$nm.

% TODO: might Howell suggest a paragraph for how the contrast ratios
% were derived?
We observed WASP-4 twice, on the night of September 11-12 (8 sets of
1000 frames each) and also on the night of September 28-29 (7 sets of
1000 frames each).
{\bf We reconstructed the images following...}
{\bf We reduced the reconstructed speckle images to contrast curves by
injecting and recovering point-sources...}
The contrast curves showed that the second night, which had better
seeing, also produces the more constraining result.
The 832$\,$nm results were also typically the most constraining, since
faint companions are typically redder.
We therefore opt for the 832$\,$nm September 28-29 results for the
remaining analysis.

@@

We considered two models for the observed transit times.  The first
model assumes a constant orbital period on a circular orbit:
\begin{align}
  t_{\rm tra}(E) &= t_0 + PE,
\end{align}
where $E$ is the epoch number and $t_0$ is a reference epoch.  The
second model assumes that the period changes at a steady rate:
\begin{align}
  t_{\rm tra}(E) &=
    t_0 + PE +
    \frac{1}{2} \frac{{\rm d}P}{{\rm d}E} E^2.
\end{align}
The free parameters are the reference epoch $t_0$, the period at the
reference epoch $P$, and the period derivative, ${\rm d}P/{\rm d}t =
(1/P) {\rm d}P/{\rm d}E$.  We defined the epoch numbers such that
$E=0$ is near the weighted average of the observed times.  This helps
to reduce the covariance between $t_0$ and $P$.  A third possible
model that we did not consider for reasons that will become apparent
is a precessing, eccentric orbit \citep[{\it
e.g.},][]{gimenez_revision_1995,patra_2017}.

@@

Figure~\ref{fig:times} shows the observed transit times, minus the
best-fit constant period model.  The best-fitting constant-period
model has $\chi^2 = 276$, 91 degrees of freedom, and $\chi^2_{\rm
red}=3.0$.  The best-fitting quadratic model has $\chi^2 = 183$, 90
degrees of freedom, and $\chi^2_{\rm red}=2.0$.  The difference in the
Bayesian information criteria (BIC) between the linear and quadratic
and models is $\Delta {\rm BIC} = 89.1$, strongly favoring the
quadratic case \citep{kass_bayes_1995}.

From the reduced $\chi^2$ values, we can surmise that neither model
entirely describes the transit data that we have collected---there
must be some additional source of signal or noise. In
\citet{bouma_wasp4b_2019}, we found that the quadratic model for the
(sparser) transit data gave $\chi^2_{\rm red}=1.0$.  The worsened
$\chi^2_{\rm red}$ could reflect underestimated statistical
uncertainties in any of our transit measurements.  It could also
reflect systematic errors in the time-systems in which the transit
measurements were recorded, though we have taken every caution against
this latter possibility.

%FIXME: might just do the example calculation?
One approach to analyzing such data would be to introduce a fudge
factor that inflates the transit measurement uncertainties, and lowers
the reduced $\chi^2$.  We do not think that such an approach is
warranted, since it would not change the result that the quadratic
model is strongly preferred.

@@

\begin{figure}[t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.5\textwidth]{f2.pdf}
	\end{center}
	\vspace{-0.7cm}
	\caption{
    {\bf Radial velocity observations of WASP-4.} The best-fit
    Keplerian orbit of WASP-4b has been subtracted.  The linear trend
    inferred from the RV data is shown with a black line, with
    $1\sigma$ errors in gray.  The trend that would be needed to
    produce the period decrease seen in transits ($\dot{P}= - 8.64 \pm
    0.89~{\rm ms}\,{\rm yr}^{-1}$) is indicated with the purple dotted
    line.  The four new RV measurements presented in this work
    increased the significance of the radial velocity trend from
    $\approx$2$\sigma$ to 15$\sigma$.
	\label{fig:rvs}
  \vspace{-0.3cm}
	}
\end{figure}

Our initial model for the radial velocity data was a single Keplerian
orbit, plus instrument offsets, jitters, and a long-term trend
\citep[][\texttt{radvel}]{fulton_radvel_2018}.  We set Gaussian priors
on the orbital period and time of inferior conjunction using the
values from Table~4 of \citetalias{bouma_wasp4b_2019}, and fixed
WASP-4b's eccentricity to zero
\citep{beerer_secondary_2011,knutson_friends_2014,bonomo_gaps_2017}.
The free parameters were the velocity semi-amplitude, the instrument
zero-points, an additive ``white noise'' instrument jitter for each
instrument, linear ($\dot{v_{\rm r}}$), and optionally second-order
($\ddot{v_{\rm r}}$) acceleration terms.

@@

WASP-4 was found to be accelerating towards our line-of-sight at high
confidence,
\begin{equation}
  \dot{v}_{\rm r} = \dot{\gamma} = 
     -0.0422^{+0.0028}_{-0.0027}
     \,{\rm m}\,{\rm s}^{-1}\,{\rm day}^{-1}.
\end{equation}
For comparison, before our new measurements, $\dot{v}_{\rm r}$ was
thought to be about five times smaller, and was only marginally
significant \citep{knutson_friends_2014,bouma_wasp4b_2019}.

The system's acceleration towards our line-of-sight causes a decrease
in the apparent orbital period.  The period derivative expected from
radial velocities is
\begin{align}
  \dot{P}_{\rm RV}
  = \frac{\dot{v}_{\rm r} P}{c}
  = -5.94 \pm 0.39~{\rm ms}\,{\rm yr}^{-1}.
\end{align}
In other words, the majority of the period decrease observed in
transits ($\dot{P}= - 8.64 \pm 0.89~{\rm ms}\,{\rm yr}^{-1}$) seems to
be explained by the acceleration of the host star.

An important consideration is whether the measured RV trend is
correlated with stellar activity.  We investigate this by analyzing
WASP-4's emission in the Ca II H \& K lines, as quantified with the
chromospheric $S$-index \citep{wright_chromospheric_2004}.  We only
examined the HIRES velocities for this step, since they are the main
source of signal for our analysis.  First, we subtracted the orbital
solution from the Keck-HIRES velocities.  Then, following
\citet{bryan_statistics_2016,bryan_excess_2019}, we calculated the
Spearman rank correlation coefficient between the $S$-index and the
orbit-subtracted velocities.  We found a correlation coefficient of
0.16. This correlation is not statistically significant; the
corresponding $p$-value is 0.65.  Furthermore, inspection of the
$S$-index timeseries showed no secular or sinusoidal trends, as would
be expected if we were observing a long-term magnetic activity cycle.
The $S$-index values are included in {\bf Table X}.  We conclude that
it is highly unlikely that the linear trend is caused by
stellar activity.

@@

\begin{figure*}[t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.78\textwidth]{f4.pdf}
	\end{center}
	\vspace{-0.6cm}
	\caption{
    {\bf Masses and semi-major axes of companions that meet
    requirements of both the radial velocity and the speckle imaging.}
    The likelihood inferred from radial velocities is shown in
    grayscale, and the region excluded from the speckle imaging is
    indicated (blue line).  The expected mass to semimajor-axis
    degeneracy is shown with a black line.
	\label{fig:mass_sma}
  \vspace{-0cm}
	}
\end{figure*}

@@

Due to the custom filters of the Zorro imager, and corresponding lack
of synthetic photometry, we further assumed that all sources had
blackbody spectra. While this is clearly false, we do not readily have
access to the planetary and stellar atmosphere models needed for the
consistent calculation with the \texttt{COND03} and \texttt{MESA}
models.  We therefore adopted the effective temperatures and
bolometric luminosities from the \citet{baraffe_evolutionary_2003} and
MIST isochrones.  Using these theoretical quantities and the
empirically-measured Zorro bandpasses, we calculated absolute
magnitudes in the 562 and 832 nm Zorro bands for stellar and planetary
mass companions.  Applying the same calculation to WASP-4 itself using
the effective temperature and bolometric luminosity from
\citetalias{bouma_wasp4b_2019}, we derived the transformation from
contrast ratio to companion mass.  The resulting limits are shown in
Figure~\ref{fig:mass_sma}.

@@

We assigned each companion a mass and semimajor axis from log-uniform
distributions within the grid cell. We drew the inclination from a
uniform distribution in $\cos i$.  For companion masses less than
$10\,M_{\rm Jup}$, we drew the eccentricity from
\citet{kipping_beta_2013}'s long-period exoplanet Beta distribution
($a=1.12$, $b=3.09$).  If the companion mass exceeded $10\,M_{\rm
Jup}$, we drew the eccentricity distribution from the power-law $p_e
\propto e^\eta$ reported by \citet{moe_mind_2017} in their
Equation~17.  We emphasize that the long-period exoplanet and
long-period binary eccentricity distributions are quite different: the
exoplanet distribution is ``bottom-heavy'', with eccentricities
preferentially close to zero. The binary star distribution is
``top-heavy'', with eccentricities closer to one.  We chose a mass
cutoff of $10\,M_{\rm Jup}$ to separate the two regimes based on the
bound observed by \citet{schlaufman_evidence_2018} between giant
planets, presumably formed through core accretion, and brown dwarfs,
presumably formed via gravitational instability.

For each simulated companion, we then drew a sample from the converged
chains of our initial model of WASP-4b, plus its linear trend. We
subtracted the planet's orbital solution, leaving RV data points with
a linear trend.  Given $(a_{\rm c}, M_{\rm c}, e_{\rm c})$ for each
simulated outer companion, and the fixed instrument offsets and
jitters from the MCMC chains, we then performed a maximum likelihood
fit for the time and argument of periastron of the outer simulated
companion.  We converted the resulting $128\times128\times512$ cube of
best-fit log-likelihood values to probabilities, and averaged over the
samples in each grid cell to derive a probability distribution in mass
and semi-major axis.  Figure~\ref{fig:mass_sma} shows the result in
grayscale.

@@

\subsection{Implications for WASP-4}
Previous potential explanations for
WASP-4b's decreasing orbital period included tidal orbital decay, orbital precession, and
light-travel time effects \citep{bouma_wasp4b_2019}.  Our new radial
velocity measurements strongly indicate that the least exotic
option---light-travel time effects---is also the most likely.  Transits show the obital period 
decreasing by 
$- 8.64 \pm 0.89~{\rm ms}\,{\rm yr}^{-1}$; the line-of-sight
acceleration observed
in radial velocities would predict a period decrease of
$-5.94 \pm 0.39~{\rm ms}\,{\rm yr}^{-1}$.
Though the quantitative agreement is not perfect, Occam's razor would
suggest that the line of sight acceleration is probably a sufficient
explanation for the apparent decrease of WASP-4b's orbital period.
% While further radial velocity observations of the system should help
% in determining the mass and semi-major axis of the companion, it seems
% unlikely that additional transit observations will yield near-term
% constraints on orbital decay.

The corresponding requirements for the companion causing the
acceleration are that it is likely either a brown-dwarf or low mass
star, and that it is between $10$-$100\,{\rm AU}$ from the host star
(Figure~\ref{fig:mass_sma}).  Given such a mass, this companion could
at one time have influenced the orbital evolution of the inner giant.
The fact that most hot Jupiters have similar massive outer
companions \citep{knutson_friends_2014,bryan_statistics_2016} is
circumstantial evidence for certain high-eccentricity formation
channels (see \citealt{dawson_johnson_2018}).
Further radial velocity monitoring should eventually reveal the
orbital parameters and minimum mass of WASP-4's companion.
